---
title: "Mortality in the United States for 2020"
author:
- name: Kieran Healy
  url: https://kieranhealy.org
  affiliation: Duke University
  affiliation_url: https://sociology.duke.edu
date: "`r Sys.Date()`"
description: |
  Looking at CDC Data on Mortality.
toc: true  
output:
  pdf_document:
    template: ~/.pandoc/templates/rmd-latex.template  
  html_document:
    highlight: tango    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
library(tidyverse)
library(janitor)

## https://kjhealy.github.io/socviz
library(socviz)

###-------------------------------------------------
### Not needed to draw the graphs
library(showtext)
showtext_opts(dpi = 300)
showtext_auto()

library(myriad)
import_myriad_semi()

theme_set(theme_myriad_semi())
###-------------------------------------------------


library(patchwork)

```

```{r}
## https://kjhealy.github.io/covdata
library(covdata)
```


```{r}

states <- nchs_wdc %>% 
  select(jurisdiction) %>% 
  unique()

```


# Overview

This document provides summary graphs of current CDC data on weekly mortality counts for the United States both nationally and by jurisdictions. CDC jurisdictions include the fifty states, along with Puerto Rico, Washington DC, and New York City. The graphs show (1) Overall weekly all-cause death counts, in comparison to the same weeks in 2015-2019, (2) COVID-19 attributed deaths (multiple cause) for 2020, (3) Weekly death counts by selected cause groups, in comparison to the same weeks in 2015-2019, (4) Percentage difference in mortality between each week in 2020 and the average of the same week between 2015 and 2019, and (5) Summary "dashboard"-style overviews of trends in overall mortality and mortality for selected causes by jurisdiction for 2020 and the previous five years. 

I also provide some figures showing cross-national comparative trends.

## Data sources

U.S. data are from the Centers for Disease Control / National Center for Health Statistics, packaged by the author in `covdata` for R: <https://kjhealy.github.io/covdata>.

The specific CDC series used are [NCHS Weekly Counts of Deaths by State and Select Causes, 2014-2019](https://data.cdc.gov/NCHS/Weekly-Counts-of-Deaths-by-State-and-Select-Causes/3yf8-kanr) and [NCHS Weekly Counts of Deaths by State and Select Causes, 2020-2021](https://data.cdc.gov/NCHS/Weekly-Counts-of-Deaths-by-State-and-Select-Causes/muzy-jte6). 

International data are from the Short Term Mortality Fluctuations series from the [Human Mortality Database](http://mortality.org/).

## Caveats 

The figures shown here are all more or less direct representations of summary mortality counts from the CDC and their difference in 2020 from counts over the preceding four or five years. They are not modeled estimates of excess deaths. I do not attempt to account for the uncertainty associated with ongoing reporting delays, e.g. by weighting the counts from jurisdictions known or likely to have delayed reporting of mortality data for 2020, such as North Carolina. Nor do I model factors such as the seasonality of counts, except implicitly by showing year-over-year comparisons, or attempt to definitively address more complex questions such as the knock-on effects of the COVID-19 pandemic on other causes of mortality. For some model-based estimates of excess mortality, as well as discussion of the challenges of producing such estimates in the context of ongoing data collection, see [this CDC resource](https://www.cdc.gov/nchs/nvss/vsrr/covid19/excess_deaths.htm) and the references therein. For the jurisdiction-by-jurisdiction figures, the rule is that the smaller the population of the area the noiser the estimates will be. This is particularly true for trends in specific causes, where the numbers can become quite small. 



# Cross-National Mortality Data

```{r}

rate_rank <- stmf %>%
  filter(sex == "b", year > 2014 & year < 2020) %>%
  group_by(country_code) %>%
  summarize(mean_rate = mean(rate_total, na.rm = TRUE)) %>% 
  mutate(rate_rank = rank(mean_rate))
  

rate_max_rank <- stmf %>%
  filter(sex == "b", year == 2020) %>%
  group_by(country_code) %>%
  summarize(covid_max = max(rate_total, na.rm = TRUE)) %>% 
  mutate(covid_max_rank = rank(covid_max))

out <- stmf %>%
  filter(sex == "b", year > 2014, year < 2021, 
         country_code %in% c("AUT", "BEL", "CHE", "DEUTNP", "DNK", "ESP", "FIN",
                             "FRATNP", "GBR_SCO", "GBRTENW", "GRC", "HUN",
                             "ITA", "LUX", "POL", "NLD", "NOR", "PRT", "SWE", "USA")) %>%
  filter(!(year == 2020 & week > 53)) %>%
  group_by(cname, year, week) %>%
  mutate(yr_ind = year %in% 2020) %>%
  slice(1) %>% 
  left_join(rate_rank, by = "country_code") %>% 
  left_join(rate_max_rank, by = "country_code") %>% 
  ggplot(aes(x = week, y = rate_total, color = yr_ind, group = year)) + 
  scale_color_manual(values = c("gray70", "firebrick"), labels = c("2015-2019", "2020")) +
  scale_x_continuous(limits = c(1, 52),  
                     breaks = c(1, seq(10, 50, 10)), 
                     labels = as.character(c(1, seq(10, 50, 10)))) + 
  facet_wrap(~ reorder(cname, rate_rank, na.rm = TRUE), ncol = 4) +
  geom_line(linewidth = 0.9) + 
  guides(color = guide_legend(override.aes = list(size = 3))) + 
  labs(x = "Week of the Year", 
       y = "Total Death Rate", 
       color = "Year",
       title = "Overall Weekly Death Rates",
       subtitle = "Comparing 2020 with 2015-2019 across selected countries. Countries are shown top\nleft to bottom right ordered from lowest to highest average mortality rate in 2015-2019.",
       caption = "Graph: @kjhealy. Data: Human Mortality Database, mortality.org") + 
  theme(legend.position = "top", 
        plot.title = element_text(size = rel(3.6)),
        plot.subtitle = element_text(size = rel(1.25)),
        strip.text = element_text(size = rel(1.4), hjust = 0),
        legend.text = element_text(size = rel(1.1)), 
        legend.title = element_text(size = rel(1.1)))


ggsave("figures/national_mortality_rates.pdf", out, height = 16, width = 10)
ggsave("figures/national_mortality_rates.png", out, height = 16, width = 10)

ggsave("figures/national_mortality_rates_wide.pdf", out, height = 10, width = 20)
ggsave("figures/national_mortality_rates_wide.png", out, height = 10, width = 20)

```

```{r, echo = FALSE, fig.align = "center", fig.show="asis"}
knitr::include_graphics("figures/national_mortality_rates.pdf")
```



# US Comparisons from Weekly Death Counts

```{r, eval=FALSE}

nchs_wdc %>% 
  filter(cause %in% c("All Cause", "Natural Causes"), jurisdiction %in% "United States") %>% 
  group_by(cause, year) %>% 
  summarize(n = sum(n)) %>% 
  pivot_wider(names_from = "cause", values_from = "n") %>% 
  mutate(delta = `All Cause` - `Natural Causes`)

```

```{r}
start_week <- 1
end_week <- 53

df_yr <- nchs_wdc %>%
  filter(year > 2014,
         year < 2021,
         week >= start_week & 
         week <= end_week) %>% 
  group_by(jurisdiction, cause, year) %>%
  summarize(period_deaths = sum(n, na.rm = TRUE)) 


baseline_deaths <- nchs_wdc %>% 
  filter(year %in% c(2015:2019),
         week >= start_week & 
         week <= end_week) %>%
  group_by(jurisdiction, year, cause) %>%
  summarize(total_n = sum(n, na.rm = TRUE)) %>%
  group_by(jurisdiction, cause) %>%
  summarize(baseline = mean(total_n, na.rm = TRUE), 
            baseline_sd = sd(total_n, na.rm = TRUE)) 

df_excess <- left_join(df_yr, baseline_deaths) %>%
  mutate(excess = period_deaths - baseline, 
         pct_excess = (excess / period_deaths)*100, 
         pct_sd = (baseline_sd/baseline)*100) %>%
  rename(deaths = period_deaths)

df_meds <- df_excess %>%
  summarize(med = median(pct_excess))
```

```{r, eval = FALSE}

df_excess %>% 
  ungroup() %>% 
  filter(jurisdiction == "United States", 
         cause == "All Cause", 
         year < 2021) %>% 
  select(jurisdiction, year, deaths)

df_excess %>% 
  ungroup() %>% 
  filter(jurisdiction == "United States", 
         cause == "All Cause", 
         year == 2020) %>% 
  select(jurisdiction, year, deaths, baseline, 
         excess, pct_excess)


```


```{r}
excess_count <- df_excess %>%
  filter(year == 2020 &
           cause %in% c("All Cause", "COVID-19 Multiple cause")) 

excess_table <- excess_count %>%
  mutate(col_cause = janitor::make_clean_names(cause)) %>%
  select(jurisdiction, col_cause, deaths:pct_excess) %>%
  group_by(jurisdiction) %>%
  select(-cause) %>%
  pivot_wider(names_from = col_cause, values_from = deaths:pct_excess) %>%
  select(-pct_excess_covid_19_multiple_cause, -excess_covid_19_multiple_cause, -baseline_covid_19_multiple_cause, 
         -baseline_sd_covid_19_multiple_cause)

colnames(excess_table) <- c("jurisdiction", "all_cause", "covid", "baseline", "baseline_sd", "excess", "pct_excess")


excess_table <- excess_table %>%
  mutate(deficit = excess - covid, 
         pct_covid = (covid / all_cause) * 100, 
         pct_deficit = (deficit / all_cause) * 100) %>%
  select(jurisdiction, all_cause, baseline, baseline_sd, excess, covid, deficit, everything()) 

excess_table %>%
  filter(jurisdiction == "United States") %>%
  pivot_longer( all_cause:pct_deficit, 
                names_to = "measure", values_to = "value")

```


```{r}

df_state_meds <- df_excess %>%
  summarize(med = median(pct_excess))

df_state_sd <- df_excess %>%
  filter(cause %nin% c("COVID-19 Underlying", "COVID-19 Multiple cause", "Other")) %>%
  group_by(jurisdiction, cause) %>% 
  slice(1) %>%
  select(cause, pct_sd) %>%
  mutate(lwr = -2*pct_sd, 
         upr = 2*pct_sd) %>%
  left_join(df_state_meds)



df_state_excess <- left_join(df_excess, df_state_sd) 

df_state_excess2020 <- df_state_excess %>% 
  select(jurisdiction,  year, cause, pct_excess) %>% 
  filter(year == 2020) %>% 
  rename(pct_excess_2020 = pct_excess) %>% 
  select(jurisdiction, cause, pct_excess_2020) %>% 
  mutate(all_excess_2020 = if_else(cause == "All Cause", pct_excess_2020, NA_real_)) 

df_state_excess2020 %>%
  ungroup() %>% 
  fill(all_excess_2020) %>% 
  group_by(jurisdiction, cause) 

df_state_excess <- left_join(df_state_excess, df_state_excess2020)

df_state_excess <- df_state_excess %>%
  ungroup() %>% 
  fill(all_excess_2020) %>% 
  group_by(jurisdiction, cause) 


```


```{r}

timestamp <-  lubridate::stamp("March 1, 1999", "%B %d, %Y")(lubridate::ymd(Sys.Date()))   

out <- df_state_excess %>% 
   filter(cause %in% c("All Cause")) %>% 
  mutate(yr_ind = ifelse(year == 2020, TRUE, FALSE)) %>%
  ggplot(aes(x = pct_excess/100, 
             y = reorder(jurisdiction, pct_excess_2020, na.rm = TRUE), 
             color = yr_ind, shape = yr_ind, group = year)) + 
  geom_linerange(mapping = aes(xmin = lwr/100, 
                               xmax = upr/100, 
                               y = reorder(jurisdiction, pct_excess_2020, na.rm = TRUE)), 
                 color = "lightblue1", alpha = 0.7, size = 3) + 
  geom_vline(xintercept = 0, color = "black") + 
  geom_jitter(size = 2, position = position_jitter(height = 0.05)) +
  scale_color_manual(values = c("gray50", "firebrick"), 
                     labels = c("2015-2019", "2020")) + 
  scale_shape_manual(values = c(16, 17),
                      labels = c("2015-2019", "2020")) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1), 
                     breaks = seq(-0.2, 0.5, 0.1), 
                     position = "top") + 
  guides(shape = guide_legend(keywidth = 0.15, override.aes = list(size = 4)), 
         color = guide_legend(keywidth = 0.15)) + 
  labs(x = "  Percent above or below the average number of deaths, 2015-2019", 
       y = NULL, 
       color = "Years",
       shape = "Years",
       title = " All-Cause Mortality in the United States:\n Comparing 2020 to 2015-2019",
       subtitle = " Provisional data. Blue bars show +/- 2 standard deviations around the 2015-2019 mean.\n Jurisdictions are ordered from highest to lowest difference from the 2015-19 baseline.",
       caption = paste0("Data: CDC. Calculations and Graph: Kieran Healy / @kjhealy\nThis graph was made on ", timestamp)) 

out <- out + theme(legend.justification = "left", 
          panel.border  = element_rect(color = "gray90", size = rel(0.9), linetype="solid", fill = NA),
          axis.text.y = element_text(face = c(rep("plain", 34), "bold", rep("plain", 19))), 
          axis.title.x = element_text(hjust = 0, size = rel(0.9)), 
          legend.text = element_text(size=rel(1.1)),
          legend.title = element_text(size=rel(1.2)),
          plot.title = element_text(size = rel(1.65)), 
          plot.subtitle = element_text(size = rel(1)),
          plot.margin = margin(1, 2, 0.5, 0.5, "cm"))

ggsave(filename = "figures/excess_all_cause_by_jurisdiction.pdf", out, height = 12, width = 8)
ggsave(filename = "figures/excess_all_cause_by_jurisdiction.png", out, height = 12, width = 8)


```

```{r, echo = FALSE, fig.align = "center", fig.show="asis", out.width="90%"}
knitr::include_graphics("figures/excess_all_cause_by_jurisdiction.pdf")
```


```{r}

states <- nchs_wdc %>% 
  select(jurisdiction) %>% 
  unique() %>%
  mutate(fname = tolower(paste0("figures/", jurisdiction, "_patch")), 
         fname = stringr::str_replace_all(fname, " ", "_"))

```

# US National and Jurisdictional Dashboards

```{r, echo = FALSE}

dat <- nchs_wdc %>%
  filter(year > 2014, year < 2021) %>%
  mutate(month_label = lubridate::month(week_ending_date, label = TRUE))

average_deaths <- nchs_wdc %>% 
  filter(year %in% c(2015:2019)) %>%
  group_by(jurisdiction, week, cause) %>%
  summarize(average_wk_deaths = mean(n, na.rm = TRUE)) 

df <- left_join(dat, average_deaths) %>%
  select(everything(), n, average_wk_deaths) %>%
  mutate(n_diff = n - average_wk_deaths, 
         pct_diff = (n_diff / n)*100) %>%
  filter(cause %nin% c("Natural Causes", "Other"))
```


```{r, echo = FALSE}
## Choose how many red-line wks
nwks <- 53

season_label <- tibble(wk_num = lubridate::epiweek(as.Date(c("2020-03-01",
                                     "2020-06-01",
                                     "2020-09-01",
                                     "2020-12-01"))),
                    season_lab = c("Spring", "Summer", "Autumn", "Winter"))


month_label <- tibble(wk_num = lubridate::epiweek(lubridate::ymd("2020-01-15") + months(0:11)),
                      month_lab = as.character(lubridate::month(lubridate::ymd("2020-01-15") + months(0:11), label = TRUE)))


order_panels <- function(st = state, ...) {
  df %>% 
  filter(year < 2021,
         jurisdiction %in% st, cause != "All Cause") %>%
  group_by(cause) %>% 
  summarize(deaths = sum(n, na.rm = TRUE), 
            .groups = "drop") %>%
  mutate(cause_rank = rank(-deaths), 
         o = order(cause_rank),
         cause_ord = factor(cause, levels = cause[o], ordered = TRUE)) %>%
  select(cause, cause_ord)
}

patch_state_count <- function(state) {

  out <- df %>% 
  filter(year < 2021, jurisdiction %in% state, cause == "All Cause") %>%
  group_by(year, week) %>% 
  mutate(yr_ind = year %in% 2020) %>%
  filter(!(year == 2020 & week > nwks)) %>%
  ggplot(aes(x = week, y = n, color = yr_ind, group = year)) + 
  geom_line(linewidth = 0.9) + 
  scale_color_manual(values = c("gray70", "firebrick"), labels = c("2015-2019", "2020")) +
  scale_x_continuous(breaks = month_label$wk_num, labels = month_label$month_lab) +      
  scale_y_continuous(labels = scales::comma) +
  labs(x = NULL, 
       y = "Total Deaths", 
       color = "Years",
       title = "Weekly recorded deaths from all causes", 
       subtitle = "Raw Counts. Provisional Data. Late 2020 counts are likely incomplete.") 
  
  out

}

patch_state_cause <- function(state, nc = 2) {

panel_ordering <- order_panels(st = state)
  
out <- df %>% 
  filter(year < 2021,
         jurisdiction == state, 
         cause %nin% c("All Cause", "COVID-19 Multiple cause", "COVID-19 Underlying")) %>%
  group_by(cause, year, week) %>% 
  summarize(deaths = sum(n, na.rm = TRUE), .groups = "drop") %>%
  mutate(yr_ind = year %in% 2020) %>%
  filter(!(year == 2020 & week > nwks)) %>%
  left_join(panel_ordering, by = "cause") %>%
  ggplot(aes(x = week, y = deaths, color = yr_ind)) + 
  geom_line(linewidth = 0.9, mapping = aes(group = year)) + 
  scale_color_manual(values = c("gray70", "firebrick"), labels = c("2015-2019", "2020")) +
  scale_x_continuous(breaks = c(1, 10, 20, 30, 40, 50), labels = as.character(c(1, 10, 20, 30, 40, 50))) + 
  scale_y_continuous(labels = scales::comma) +
  facet_wrap(~ cause_ord, ncol = nc, labeller = label_wrap_gen(25)) + 
  labs(x = "Week of the Year", 
       y = "Total Deaths", 
       color = "Years",
       title = "Weekly deaths from selected causes", 
       subtitle = "Panels ordered by number of deaths. Raw Counts.") 

out
}

patch_state_percent <- function(state, nc = 2){
  
  panel_ordering <- order_panels(st = state)

  out <- df %>%
  filter(year < 2021) %>%   
  filter(jurisdiction %in% state, 
         year == 2020, 
         cause %nin% c("All Cause", "COVID-19 Multiple cause", 
                                                "COVID-19 Underlying"), !is.na(pct_diff)) %>%
  group_by(week) %>% 
  filter(!(week > nwks)) %>%
  mutate(ov_un = pct_diff > 0) %>%
  left_join(panel_ordering, by = "cause") %>%
  ggplot(aes(x = week, y = pct_diff/100, fill = ov_un)) + 
  geom_col() + 
  scale_x_continuous(breaks = c(1, seq(10, nwks, 10)), labels = as.character(c(1, seq(10, nwks, 10)))) + 
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("gray40", "firebrick")) +
  guides(fill = "none") + 
  facet_wrap(~ cause_ord, ncol = nc, labeller = label_wrap_gen(25)) + 
  labs(x = "Week of the Year", 
       y = "Percent", 
       title = "Percent difference from 2015-2019 average") 

  out
}

patch_state_covid <- function(state) {

  out <- df %>% 
  filter(year < 2021) %>%   
  filter(jurisdiction %in% state, cause %in% c("COVID-19 Multiple cause")) %>%
  group_by(year, week) %>% 
  mutate(yr_ind = year %in% 2020) %>%
  filter(year == 2020) %>%
  ggplot(aes(x = week, y = n, group = year)) + 
  geom_col(fill = "gray30") + 
  scale_x_continuous(breaks = c(1, 10, 20, 30, 40, 50), 
                     labels = as.character(c(1, 10, 20, 30, 40, 50)), 
                     limits = c(1, 52)) + 
  scale_y_continuous(labels = scales::comma) + 
  labs(x = "Week of the Year", 
       y = "Total Deaths", 
       color = "Years",
       subtitle = "Raw counts.",
       title = "Weekly deaths recorded as COVID-19 (Multiple cause)") 
  
  out

}

make_patchplot <- function(state){
  
if(state == "New York")  {
  state_title <- paste(state, "(Excluding New York City)")
} else 
  {
  state_title <- state
}

timestamp <-  lubridate::stamp("March 1, 1999", "%B %d, %Y")(lubridate::ymd(Sys.Date()))   
  


(patch_state_count(state) + theme(plot.margin = unit(c(5,0,0,0), "pt"))) / patch_state_covid(state) / (patch_state_cause(state) + (patch_state_percent(state))) +  
    plot_layout(heights = c(2, 0.5, 4), guides = 'collect') + 
  plot_annotation(
  title = state_title,
  caption = paste0("Graph: @kjhealy Data: CDC. This graph was made on ", timestamp, "."), 
  theme = theme(plot.title = element_text(size = rel(2), hjust = 0, face = "plain")))
}
```

```{r draw-everything, eval = TRUE}

out_patch <- states %>% 
  mutate(patch_plot = map(jurisdiction, make_patchplot))
```

```{r walk-everything, eval = TRUE}

walk2(paste0(out_patch$fname, ".png"), 
     out_patch$patch_plot, ggsave, height = 18, width = 12)

walk2(paste0(out_patch$fname, ".pdf"), 
     out_patch$patch_plot, ggsave, height = 18, width = 12)




```

```{r}

us_average_deaths <- nchs_wdc %>% 
  filter(year %in% c(2015:2019)) %>%
  group_by(week, cause) %>%
  summarize(average_wk_deaths = mean(n, na.rm = TRUE)) %>%
  ungroup()

us_total <- nchs_wdc %>%
  group_by(year, week, cause) %>%
  summarize(n = sum(n, na.rm = TRUE)) %>%
  ungroup()

us_all <- left_join(us_total, us_average_deaths) %>%
  filter(year %in% c(2015:2019)) %>%
  select(everything(), n, average_wk_deaths) %>%
  mutate(n_diff = n - average_wk_deaths, 
         pct_diff = (n_diff / n)*100)
  
us <- make_patchplot("United States")

ggsave("figures/usa_patch.pdf", us, height = 18, width = 12)
ggsave("figures/usa_patch.png", us, height = 18, width = 12)

```

```{r, echo = FALSE, fig.align = "center", fig.show="asis", out.width="90%"}
knitr::include_graphics("figures/usa_patch.pdf")
```


```{r, echo = FALSE, fig.align = "center", fig.show="asis", out.width="90%"}
patchplots <- list.files("figures", pattern = ".*patch.pdf", full.names = TRUE)
knitr::include_graphics(patchplots)
```







